Quantum Hoare Type Theory

### 2020-04-29 ms-paper/ms-paper.tex
Switch to listings package with Haskell language

\usepackage[braket]{qcircuit}
\usepackage{mathpartir}
\usepackage{mathtools}
\usepackage{xspace}
\usepackage{fancyvrb}
\usepackage{cleveref}
\crefformat{section}{\S#2#1#3}

\usepackage{listings}
\usepackage{lstcoq}

\definecolor{ltblue}{rgb}{0,0.4,0.4}
\definecolor{dkblue}{rgb}{0,0.1,0.6}
\definecolor{dkgreen}{rgb}{0,0.35,0}
\definecolor{dkviolet}{rgb}{0.3,0,0.5}
\definecolor{dkred}{rgb}{0.5,0,0}

\lstset{language=coq, basicstyle=\ttfamily\scriptsize, columns=flexible,
	keepspaces=true}

@@

Here we discuss the relevant bits from the QIO interface. QIO provides a type for referring to qubits, \texttt{Qbit}; a type for unitary operations, \texttt{U}; and, the \texttt{QIO} type operator which is a monad indexed by the result of the quantum computation. The primitive quantum operations defined using the QIO monad below (Haskell syntax) are:
\begin{verbatim}
mkQbit : Bool -> QIO Qbit       -- initialization
applyU : U -> QIO ()            -- apply a unitary to quantum state
measQbit : Qbit -> QIO Bool     -- measurement
\end{verbatim}

@@

\begin{verbatim}
rot :: Qbit -> ((Bool, Bool) -> C) -> U
cond :: Qbit -> (Bool -> U) -> U
\end{verbatim}

@@

\begin{verbatim}
ifQ :: Qbit -> U -> U
ifQ q u = cond q (\x -> if x then u else mempty)
\end{verbatim}

@@

\subsubsection*{Hello Quantum World}
\begin{verbatim}
hqw : {emp} r : Bool {emp /\ Id(r, false)}
    = do q <= mkQbit false;
         measQbit q
\end{verbatim}

@@

\subsubsection*{Coin Toss}
\begin{verbatim}
rnd : {emp} r : Bool {emp}
    = do q <= mkQbit false;
         applyU (H q);
         measQbit q
\end{verbatim}

@@

\begin{Verbatim}[commandchars=\$\[\]]
testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}
              -- P0: emp
         = do qa <= mkQbit false;
              -- P1: P0 $ccc (qa $cmt state(false))
              applyU (H qa);
              -- P2: P1 $ccc (qa $cmt state({(0.5, false), (0.5, true)}))
              qb <= mkQbit false;
              -- P3: P2 $ccc (qa $cmt state(false))
              applyU (ifQ qa (X qb));
              -- P4: P3 $ccc ((qa, qb) $cmt state({(0.5, (false, false),
              --                              (0.5, (true, true))}))
                (measQbit qa, measQbit qb)
              -- P5: P4 $ccc ((qa $cmt -) $cmm emp) $ccc ((qb $cmt -) $cmm emp)
\end{Verbatim}
Finally, the \texttt{consq} rule should ensure that the assertion P5 can lead to the specified postcondition in the type of the program.

@@

%\begin{lstlisting}

@@

%\end{lstlisting}

@@

%		\begin{lstlisting}

@@

%		\end{lstlisting}

@@

\subsection{Hadamard basis states}
\begin{Verbatim}[commandchars=\$\[\]]
qplus : {emp} r : Qbit {Id(r, $cket[+])}
      = do q <= mkQbit false;
           applyU (H q);
           return q

qminus : {emp} r : Qbit {Id(r, $cket[-])}
       = do q <= mkQbit true;
            applyU (H q);
            return q
\end{Verbatim}

\subsection{Creating entanglement}
\begin{Verbatim}[commandchars=\$\[\]]
share : $cpi a : Qbit.
        {a $cin {$cket[+], $cket[-]}}   -- a $cin {b, c} is short for Id(a,b) \/ Id(a,c)
           b : Qbit
        {Id(a, b) /\ a $cin {$cket[0], $cket[1]}}
      = \a.do b <= mkQbit false;
              applyU (ifQ a (X b));
              return b
\end{Verbatim}

\subsection{Bell pair}
\begin{Verbatim}[commandchars=\$\[\]]
bell : {emp} (a, b) : (Qbit, Qbit) {Id(a, b) /\ a $cin {$cket[0], $cket[1]}}
     = do qa <- qplus;
          qb <- share qa;
          return (qa, qb)
\end{Verbatim}

\subsection{Testing modular Bell pair}
\begin{Verbatim}[commandchars=\$\[\]]
testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}
         = do (qa, qb) <- bell;
              (measQbit qa, measQbit qb)
\end{Verbatim}

\section{Teleportation}
\label{app:teleport}
% \todo{TODO: Refine assertions}
\subsection{Alice's circuit}
\begin{Verbatim}[commandchars=\$\[\]]
alice : $cpi a : Qbit. $cpi e : Qbit
        {(Id(a, -) /\ entangled(e)}
           r : (Bool, Bool)
        {emp}
      = \a.\e.do applyU (ifQ a (X e));
                 applyU (H a);
                 (measQbit a, measQbit e)
\end{Verbatim}

\subsection{Bob's circuit}
\begin{Verbatim}[commandchars=\$\[\]]
bob : $cpi m1 m2 : Bool. $cpi e : Qbit
      {entangled(e)}
         r : Qbit
      {(Id(r, -)}
      = \m1.\m2.\e.do if m1 then applyU (Z e) else ();
                      if m2 then applyU (X e) else ();
                      return e
\end{Verbatim}

\subsection{Teleport}
\begin{Verbatim}[commandchars=\$\[\]]
teleport : $cpi q : Qbit. x : Pure.
      {Id(q, x)}
         r : Qbit
      {Id(r, x)}
      = \q.do (a, b) <- bell;
              (m1, m2) <- alice q a;
              tq <- bob m1 m2 b;
              return tq
\end{Verbatim}

>>>>

\usepackage[braket]{qcircuit}
\usepackage{mathpartir}
\usepackage{mathtools}
\usepackage{xspace}
\usepackage{cleveref}
\crefformat{section}{\S#2#1#3}

\usepackage{xcolor}
\definecolor{ltblue}{rgb}{0,0.4,0.4}
\definecolor{dkblue}{rgb}{0,0,1}
\definecolor{dkgreen}{rgb}{0,0.35,0}
\definecolor{dkviolet}{rgb}{0.3,0,0.5}
\definecolor{dkred}{rgb}{0.5,0,0}

\usepackage{listings}
\lstloadlanguages{Haskell}
\lstdefinelanguage{QHaskell}{%
	language     = Haskell,
	morekeywords = {Qbit, U, C},
}
\lstnewenvironment{QCode}
{\lstset{language=QHaskell}%
	\csname lst@SetFirstLabel\endcsname}
{\csname lst@SaveFirstLabel\endcsname}
\lstset{
	basicstyle=\small\ttfamily,
	flexiblecolumns=false,
	basewidth={0.5em,0.45em},
	commentstyle=\color{dkgreen},
	keywordstyle=\bfseries\color{dkblue},
	stringstyle=\color{dkviolet},
	literate={+}{{$+$}}1 {/}{{$/$}}1 {*}{{$*$}}1 {=}{{$=$}}1
	{>}{{$>$}}1 {<}{{$<$}}1 {\\}{{$\lambda$}}1
	{\\\\}{{\char`\\\char`\\}}1
	{->}{{$\rightarrow$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
	{<=}{{$\Leftarrow$}}2 {=>}{{$\Rightarrow$}}2
	{\ .}{{$\circ$}}2 {\ .\ }{{$\circ$}}2
	{>>}{{>>}}2 {>>=}{{>>=}}2
	{|}{{$\mid$}}1
	{/\\}{{$\wedge$}}1 {\\in}{{$\in$}}1
}

@@

Here we discuss the relevant bits from the QIO interface. QIO provides a type for referring to qubits, \texttt{Qbit}; a type for unitary operations, \texttt{U}; and, the \texttt{QIO} type operator which is a monad indexed by the result of the quantum computation. The primitive quantum operations defined using the QIO monad below (Haskell syntax) are:
\begin{QCode}
mkQbit : Bool -> QIO Qbit       -- initialization
applyU : U -> QIO ()            -- apply a unitary to quantum state
measQbit : Qbit -> QIO Bool     -- measurement
\end{QCode}

@@

\begin{QCode}
rot :: Qbit -> ((Bool, Bool) -> C) -> U
cond :: Qbit -> (Bool -> U) -> U
\end{QCode}

@@

\begin{QCode}
ifQ :: Qbit -> U -> U
ifQ q u = cond q (\x -> if x then u else mempty)
\end{QCode}

@@

\subsubsection*{Hello Quantum World}
\begin{QCode}
hqw : {emp} r : Bool {emp /\ Id(r, false)}
    = do q <= mkQbit false;
         measQbit q
\end{QCode}

@@

\subsubsection*{Coin Toss}
\begin{QCode}
rnd : {emp} r : Bool {emp}
    = do q <= mkQbit false;
         applyU (H q);
         measQbit q
\end{QCode}

@@

\begin{QCode}
testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}
              -- P0: emp
         = do qa <= mkQbit false;
              -- P1: P0 $ccc (qa $cmt state(false))
              applyU (H qa);
              -- P2: P1 $ccc (qa $cmt state({(0.5, false), (0.5, true)}))
              qb <= mkQbit false;
              -- P3: P2 $ccc (qa $cmt state(false))
              applyU (ifQ qa (X qb));
              -- P4: P3 $ccc ((qa, qb) $cmt state({(0.5, (false, false),
              --                              (0.5, (true, true))}))
                (measQbit qa, measQbit qb)
              -- P5: P4 $ccc ((qa $cmt -) $cmm emp) $ccc ((qb $cmt -) $cmm emp)
\end{QCode}
Finally, the \texttt{consq} rule should ensure that the assertion P5 can lead to the specified postcondition in the type of the program.

@@

%\begin{QCode}

@@

%\end{QCode}

@@

%		\begin{QCode}

@@

%		\end{QCode}

@@

\subsection{Hadamard basis states}
\begin{QCode}
qplus : {emp} r : Qbit {Id(r, $cket[+])}
      = do q <= mkQbit false;
           applyU (H q);
           return q

qminus : {emp} r : Qbit {Id(r, $cket[-])}
       = do q <= mkQbit true;
            applyU (H q);
            return q
\end{QCode}

\subsection{Creating entanglement}
\begin{QCode}
share : $cpi a : Qbit.
        {a $cin {$cket[+], $cket[-]}}   -- a $cin {b, c} is short for Id(a,b) \/ Id(a,c)
           b : Qbit
        {Id(a, b) /\ a $cin {$cket[0], $cket[1]}}
      = \a.do b <= mkQbit false;
              applyU (ifQ a (X b));
              return b
\end{QCode}

\subsection{Bell pair}
\begin{QCode}
bell : {emp} (a, b) : (Qbit, Qbit) {Id(a, b) /\ a $cin {$cket[0], $cket[1]}}
     = do qa <- qplus;
          qb <- share qa;
          return (qa, qb)
\end{QCode}

\subsection{Testing modular Bell pair}
\begin{QCode}
testBell : {emp} (a, b) : (Bool, Bool) {emp /\ Id(a, b)}
         = do (qa, qb) <- bell;
              (measQbit qa, measQbit qb)
\end{QCode}

\section{Teleportation}
\label{app:teleport}
% \todo{TODO: Refine assertions}
\subsection{Alice's circuit}
\begin{QCode}
alice : $cpi a : Qbit. $cpi e : Qbit
        {(Id(a, -) /\ entangled(e)}
           r : (Bool, Bool)
        {emp}
      = \a.\e.do applyU (ifQ a (X e));
                 applyU (H a);
                 (measQbit a, measQbit e)
\end{QCode}

\subsection{Bob's circuit}
\begin{QCode}
bob : $cpi m1 m2 : Bool. $cpi e : Qbit
      {entangled(e)}
         r : Qbit
      {(Id(r, -)}
      = \m1.\m2.\e.do if m1 then applyU (Z e) else ();
                      if m2 then applyU (X e) else ();
                      return e
\end{QCode}

\subsection{Teleport}
\begin{QCode}
teleport : $cpi q : Qbit. x : Pure.
      {Id(q, x)}
         r : Qbit
      {Id(r, x)}
      = \q.do (a, b) <- bell;
              (m1, m2) <- alice q a;
              tq <- bob m1 m2 b;
              return tq
\end{QCode}
